Appendix S2. Evolutionary drivers of species and community dynamics

### 2024-07-26 manuscript/Appendix_S2_v01.Rmd
Began Species Evol, added new vignette with evol hypotheses

---
title: "Appendix S1. Linear regression models and non-linear population dynamics"
author:
  - Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]
  - Ruben J. Hermann^[University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen, Germany, ruben.hermann@uni-due.de]
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    extra_dependencies: ["float"]
  bookdown::html_document2: default
  word_document:
    reference_docx: docx_template.docx
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Author Pantel and Hermann}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
link-citations: yes
linkcolor: blue
csl: the-american-naturalist.csl
bibliography: refs.bib
---

@@

# One species, logistic growth
Population growth over time in a single species is first modelled using a Beverton-Holt (discrete-time, logistic) model (@Beverton1957), using an intra-specific competition coefficient for density-dependent growth (@Hart2013) - thus $\alpha$ = $1/K$.

$$ N_{i,t+1} = \frac{r_i N_{i,t}}{1 + \alpha_{ii} N_{i,t}}$$
Note that in this model, the system is at equilibrium when $N_{i,t+1} = N_{i,t}$, and therefore: $$ N^* = N^*\frac{r_i}{1 + \alpha_{ii} N^*} $$

$$ 1 = \frac{r_i}{1 + \alpha_{ii} N^*} $$

$$ N^* = \frac{r_i - 1}{\alpha_{ii}} $$

## Population dynamics simulation
In the metacommunity simulation in the main text, a species resides in a site with an initial population size $N_{i,0} \sim Pois(10)$, a growth rate $r_i$ that depends on the local environmental value $E_k$ and the species trait $x_i$, and a fixed intra-specific competition coefficient of $\alpha_{ii}$ = 0.00125. We simulate population growth here:

```{r}
set.seed(42)
# Simulate initial species population growth
N1.0 <- rpois(1,10)
r1.0 <- 1.67
alpha.11 <- 0.00125
# model function
disc_log <- function(r, N0, alpha) {
    Nt1 <- (r*N0) / (1+alpha*N0)
    return(Nt1)
}
# Simulation of model for t time steps
t <- 30
N <- rep(NA, t)
N[1] <- N1.0
for (i in 2:t) {
  N[i] <- disc_log(r=r1.0, N0=N[i-1], alpha=alpha.11)
}
```
```{r figs1, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125."}
# Plot simulation: ggplot
dat <- as.data.frame(N)
dat$time <- 1:t
ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),
    linetype = "dashed", color = "gray")
```

## Linear statistical model

We fit the population time series data to a first-order auto-regressive model to predict $N_{t+1}$ as a function of $N_t$, and compare that to a linear regression. We use ln-N after @Ives1995, as also discussed in @Certain2018 and @Olivenca2021.

$$ N_{t+1} = \beta_0 + \beta_1 N_t + \epsilon_t $$
```{r,warning=FALSE}
# Fit the model
m.1.ar <- arima(x = log(N), order = c(1,0,0), include.mean = T, method="CSS")
m.1.lm <- lm(log(dat$N[2:t]) ~ log(dat$N[1:(t-1)]))
#plotting the series along with the fitted values
m.1.ar.fit <- log(N) - residuals(m.1.ar)
m.1.lm.fit <- log(dat$N[2:t]) - m.1.lm$resid
dat$ar1.fit <- m.1.ar.fit
dat$lm.fit <- NA
dat$lm.fit[2:t] <- m.1.lm.fit
```
```{r figs2, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line).",warning=FALSE}
colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
ggplot(data=dat,aes(x=time)) +
  geom_point(aes(y=log(N),color="data")) +
  geom_point(aes(y=ar1.fit, color = "AR1 fit"), alpha=0.5) +
  geom_point(aes(y=lm.fit, color = "lm fit"), alpha=0.5) +
  geom_line(aes(y=log(N), color="data"), linewidth=.4) +
  geom_line(aes(y=ar1.fit, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
  geom_line(aes(y=lm.fit, color = "lm fit"), alpha=0.5, linewidth=.4) +
  theme_bw() +
  xlab("time") +
  ylab("N") +
  labs(color="Legend") +
  scale_color_manual(values = colors)
```

The linear model is a good fit, and $N_{t+1}$ and $N_t$ are well-represented by a linear function:

```{r figs3, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$. "}
dN <- rep(NA, (t-1))
for (i in 1:(t-1)) {
  dN[i] <- log(N[i+1])-log(N[i])
}
d.dN <- as.data.frame(dN)
d.dN$Nt <- log(dat$N[1:(t-1)])
d.dN$Nt_plus_1 <- log(dat$N[2:t])
ggplot2::ggplot(d.dN, aes(x=Nt,y=Nt_plus_1)) + geom_point()
```

We also examine density dependence by plotting $\Delta N = N_{t+1} - N_t$ vs. $N_t$:

```{r figs4, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs4} Change in population size from one time step to the next $N_{t+1}$ as a function of $N_{t+1}$"}
d.dN <- as.data.frame(dN)
d.dN$Nt <- log(dat$N[1:(t-1)])
d.dN$Nt_plus_1 <- log(dat$N[2:t])
ggplot2::ggplot(d.dN, aes(Nt, dN)) + geom_point()
```
## Bayesian linear statistical model: HMSC
We can estimate the same model parameters using HMSC:

```{r}
# prepare data in HMSC format
Y <- as.matrix(log(dat$N[2:t]))
XData <- data.frame(x=log(dat$N[1:(t-1)]))
m.1.hmsc <- Hmsc(Y=Y,XData=XData,XFormula=~x)
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 1000
transient <- 500*thin
verbose <- 500*thin
# sample MCMC
m.1.sample <- sampleMcmc(m.1.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
```

```{r}
m.post.hmsc <- convertToCodaObject(m.1.sample)
summary(m.post.hmsc$Beta)
plot(m.post.hmsc$Beta)
```

These estimates match well with those from the AR1 and linear model:

```{r}
# AR1 coefficients (recall that the intercept is the term below multipled by 1 - phi1)
m.1.ar$coef
m.1.ar$coef[2]*(1-m.1.ar$coef[1])
# linear model
summary(m.1.lm)$coefficients[1:2,1:2]
# Bayesian estimates
summary(m.post.hmsc$Beta)$statistics[1:2,1:2]
```

```{r figs5, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
Gradient <- constructGradient(m.1.sample,focalVariable="x",ngrid=29)
predY <- predict(m.1.sample,Gradient=Gradient,expected=TRUE)
#preds <-computePredictedValues(m.1.sample)
plotGradient(m.1.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="N_t",ylab="predicted N_t+1")
```

```{r}
lm_dat <- data.frame(cbind(log(dat$N[2:t]),log(dat$N[1:(t-1)])))
colnames(lm_dat) <- c("Nt1","Nt")
ggplot(lm_dat,aes(Nt,Nt1)) +
  stat_summary(fun.data= mean_cl_normal) + 
  geom_smooth(method='lm')
```

## Conclusions
In this example, a first-order auto-regressive model works well, bypassing the need to estimate logistic growth parameters $r_i$ and $\alpha_{ii}$. The density-dependence dynamics ($\Delta N \sim f(N_t)$) show an overall declining trend over time. The Bayesian estimation implemented in HMSC gives good parameter estimates.

# One species, logistic growth, environmental covariate
We now consider using a linear model to analyze population growth when the species growth rate is impacted by a single environmental covariate.

## Growth depends on environment

First we add environment-dependent growth rate. The growth rate $r_i$ becomes:

$$ r_i = \hat{W}e^{-(E-x_{i,t})^2}$$

Here, $\hat{W}$ is the maximal population growth rate (set to `r r1.0` as above), $E$ is the local environmental trait optimum value, and $x_{i,t}$ is species *i* trait value at time *t*. We see that if $E = x_{i,t}$ then the growth rate is at the value $r$ = `r r1.0`. Here, we begin with $E = x_{i,t} = 0.8$, then simulate the environment $E$ value fluctuating randomly over time, and finally use a linear model to fit $E$ as a covariate.

```{r}
# Simulate initial species population growth with environment fluctuations
N1.0 <- 10
r1.0 <- 1.67
alpha.11 <- 0.00125
E.0 <- 0.8
x1.0 <- 0.8
# model function
disc_log_E <- function(r, N0, alpha, E, x) {
    Nt1 <- ((r*exp(-(E-x)^2))*N0) / (1+alpha*N0)
    return(Nt1)
}
# Simulation of model for t time steps
t <- 40
N <- rep(NA, t)
N[1] <- N1.0
E <- rep(NA, t)
E[1] <- E.0
for (i in 2:t) {
  N[i] <- disc_log_E(r=r1.0, N0=N[i-1], alpha=alpha.11, E=E[i-1], x=x1.0)
  E[i] <- E[i-1] + rnorm(1,0,.1)
}
```
```{r figs6, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125. Relationship between E and Nt is also shown."}
# Plot simulation: ggplot
dat <- as.data.frame(cbind(N,E))
dat$time <- 1:t
p1 <- ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
p2 <- ggplot2::ggplot(dat, aes(time, E)) + geom_point()
p1 + p2
```

## Linear statistical model with environmental covariate

We now include environment $E$ as a covariate in the linear model:

$$ N_{t} = \beta_0 + \beta_1 N_{t-1} + \beta_2 E_{t-1}+ \epsilon_t $$

```{r,warning=FALSE}
# Fit the model
m.2.ar <- arima(x = log(N), order = c(1,0,0), include.mean = T, method="CSS", xreg=E)
m.2.lm <- lm(log(dat$N[2:t]) ~ log(dat$N[1:(t-1)]) + log(E[1:(t-1)]))
#plotting the series along with the fitted values
m.2.ar.fit <- log(N) - residuals(m.2.ar)
m.2.lm.fit <- log(dat$N[2:t]) - m.2.lm$resid
dat$ar2.fit <- m.2.ar.fit
dat$lm2.fit <- NA
dat$lm2.fit[2:t] <- m.2.lm.fit
```
```{r figs7, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line).",warning=FALSE}
colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
ggplot(data=dat,aes(x=time)) +
  geom_point(aes(y=log(N),color="data")) +
  geom_point(aes(y=ar2.fit, color = "AR1 fit"), alpha=0.5) +
  geom_point(aes(y=lm2.fit, color = "lm fit"), alpha=0.5) +
  geom_line(aes(y=log(N), color="data"), linewidth=.4) +
  geom_line(aes(y=ar2.fit, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
  geom_line(aes(y=lm2.fit, color = "lm fit"), alpha=0.5, linewidth=.4) +
  theme_bw() +
  xlab("time") +
  ylab("N") +
  labs(color="Legend") +
  scale_color_manual(values = colors)
```

```{r figs8, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$. "}
dN <- rep(NA, (t-1))
for (i in 1:(t-1)) {
  dN[i] <- log(N[i+1])-log(N[i])
}
d.dN <- as.data.frame(dN)
d.dN$Nt <- log(dat$N[1:(t-1)])
d.dN$Nt_plus_1 <- log(dat$N[2:t])
p1 <- ggplot2::ggplot(d.dN, aes(x=Nt,y=Nt_plus_1)) + geom_point()

d.dN <- as.data.frame(dN)
d.dN$Nt <- log(dat$N[1:(t-1)])
d.dN$Nt_plus_1 <- log(dat$N[2:t])
d.dN$E <- log(dat$E[2:t])
p2 <- ggplot2::ggplot(d.dN, aes(x=E,y=Nt_plus_1)) + geom_point()
p3 <- ggplot2::ggplot(d.dN, aes(Nt, dN)) + geom_point()
p4 <- ggplot2::ggplot(d.dN, aes(E, dN)) + geom_point()

p1+p2+p3+p4
```

The linear model is a good fit when including the environmental covariate. $N_{t+1}$ and $N_t$ can still be captured by a linear relationship. However we see that the relationship between $N_{t+1}$ and $E_t$ is non-linear. This tells us that the lm is good for predictions, but not for inference (for capturing well the relationsip between the predictor and response variable). The use of linear relationships in JSDMs is discussed in [@Ingram2020], and in many applications (e.g. [@Erickson2023]) quadratic terms are used, which create bell-shaped response curves that may better match species with optimal niches (as opposed to linear, monotonically increasing relationships between population size and environmental predictors). We thus include a quadratic term for $E_t$ to provide a better fit to the data.

```{r}
df <- data.frame(cbind(log(dat$N[2:t]),log(dat$N[1:(t-1)]),E[1:(t-1)],E[1:(t-1)]^2))
colnames(df) <- c("Nt1","Nt","E","Esq")
m.2.lm <- lm(Nt1 ~ Nt + E + Esq,data=df)
```

```{r figs9, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$. "}
jtools::effect_plot(m.2.lm,pred=E,plot.points = TRUE)
ggplot(data = df, aes(x = E, y = Nt1)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2))
```

## Bayesian linear statistical model: HMSC
We can estimate the same model parameters using HMSC:

```{r}
# prepare data in HMSC format
Y <- as.matrix(log(dat$N[2:t]))
XData <- df
m.2.hmsc <- Hmsc(Y=Y,XData=XData,XFormula=~Nt+E+Esq)
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 1000
transient <- 500*thin
verbose <- 500*thin
# sample MCMC
m.2.sample <- sampleMcmc(m.2.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
```

```{r}
m2.post.hmsc <- convertToCodaObject(m.2.sample)
summary(m2.post.hmsc$Beta)
bayesplot::mcmc_trace(m2.post.hmsc$Beta)
bayesplot::mcmc_areas(m2.post.hmsc$Beta,area_method = c("equal height"))
```

These estimates match well with those from the AR1 and linear model:

```{r}
# AR1 coefficients (recall that the intercept is the term below multipled by 1 - phi1)
m.2.ar$coef
m.2.ar$coef[2]*(1-m.2.ar$coef[1])
# linear model
summary(m.2.lm)$coefficients[1:4,1:2]
# Bayesian estimates
summary(m2.post.hmsc$Beta)$statistics[1:4,1:2]
```

We recall that the interpretation of the coefficients in an arimaX (arima with covariates) model is difficult. They do not give the impact on $N_t$ per unit increase in X as in a regression. So we do not interpret the causation implied by the coefficient in the arimaX model. In the regression model, we can see that $E$ has a positive impact on $N_t$.

```{r figs10, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
Gradient <- constructGradient(m.2.sample,focalVariable="E",non.focalVariables=list(Nt=list(2),Esq=list(2)),ngrid=39)
# Esq is manually constructed as gradient-produced E^2
Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
predY <- predict(m.2.sample,XData=Gradient$XDataNew,expected=TRUE)
plotGradient(m.2.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="E_t",ylab="predicted N_t+1")
```

## Conclusions
In this example, the linear regression again works well to describe the impact of $E_t$ for $N_t$ when using the quadratic formulation. The arimaX model works well for fitting and subsequent prediction, but less well for inference about the impacts of $E$. From the quadratic regression terms for $E$, we correctly see that the population size is maximal at the species trait value and decreases away from that value. We will continue to use log-transformed abundance and now introduce quadratic terms for the environmental parameter.

# Two species, logistic growth, competition
Here we investigate how a second species impacts the inference we can make from linear models.

## Interspecific competition
The growth equation for each species now becomes:

$$ N_{i,t+1} = \frac{r_i N_{i,t}}{1 + \alpha_{ii} N_{i,t} + \alpha_{ij} N_{j,t}}$$
We use a distinct growth rate for the 2nd species and introduce the interspecific interaction coefficient $\alpha_{ij}$.

```{r}
# Simulate initial species population growth with a second species present
N1.0 <- 10
N2.0 <- 10
r1.0 <- 1.67
r2.0 <- 1.7
alpha.11 <- 0.00125
alpha.22 <- 0.00125
alpha.12 <- 0.008
alpha.21 <- 0.008725
# model function
disc_LV_comp <- function(r1,r2,N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21) {
    Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
    Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
    return(c(Nt1,Nt2))
}
# Simulation of model for t time steps
t <- 40
N <- array(NA,dim=c(t,2))
N <- as.data.frame(N)
colnames(N) <- c("N1","N2")
N$N1[1] <- N1.0
N$N2[1] <- N2.0
for (i in 2:t) {
  res <- disc_LV_comp(r1=r1.0,r2=r2.0,N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21)
  N$N1[i] <- res[1]
  N$N2[i] <- res[2]
}
```
```{r figs11, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_1$ = 1.67, $r_2$ = 1.7, $N_{i,0}$ = 10, $\\alpha_{ii}$ = 0.00125, $\\alpha_{12}$ = 0.008, and $\\alpha_{21}$ = 0.008725."}
# Plot simulation: ggplot
N$time <- 1:t
dat <- melt(N, id.vars="time")
ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
```
## Linear statistical model with covariate for both species

We fit each species' population time series to an arimaX and lm with population size of the others species as a covariate.

$$ N_{1,t} = \beta_0 + \beta_1 N_{1,t-1} + \beta_2 N_{2,t-1}+ \epsilon_t $$

```{r figs12, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with competition."}
# Plot simulation: ggplot
dat <- as.data.frame(cbind(N$N1,N$N2))
colnames(dat) <- c("N1","N2")
dat$time <- 1:t
p1 <- ggplot2::ggplot(dat, aes(time, N1)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
p2 <- ggplot2::ggplot(dat, aes(time, N2)) + geom_point()
p1 + p2
```

```{r,warning=FALSE}
## Fit the model for Species 1
m.3.ar.n1 <- arima(x = log(N$N1), order = c(1,0,0), include.mean = T, method="CSS", xreg=N$N2)
m.3.lm.n1 <- lm(log(dat$N1[2:t]) ~ log(dat$N1[1:(t-1)]) + log(dat$N2[1:(t-1)]))
#plotting the series along with the fitted values
m.3.ar.fit.n1 <- log(N$N1) - residuals(m.3.ar.n1)
m.3.lm.fit.n1 <- log(dat$N1[2:t]) - m.3.lm.n1$resid
dat$ar3.fit.n1 <- m.3.ar.fit.n1
dat$lm3.fit.n1 <- NA
dat$lm3.fit.n1[2:t] <- m.3.lm.fit.n1
## Species 2
m.3.ar.n2 <- arima(x = log(N$N2), order = c(1,0,0), include.mean = T, method="CSS", xreg=N$N1)
m.3.lm.n2 <- lm(log(dat$N2[2:t]) ~ log(dat$N2[1:(t-1)]) + log(dat$N1[1:(t-1)]))
#plotting the series along with the fitted values
m.3.ar.fit.n2 <- log(N$N2) - residuals(m.3.ar.n2)
m.3.lm.fit.n2 <- log(dat$N2[2:t]) - m.3.lm.n2$resid
dat$ar3.fit.n2 <- m.3.ar.fit.n2
dat$lm3.fit.n2 <- NA
dat$lm3.fit.n2[2:t] <- m.3.lm.fit.n2
```

```{r figs13, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (gray line) and from a linear model (orange line).",warning=FALSE}
colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
p1 <- ggplot(data=dat,aes(x=time)) +
  geom_point(aes(y=log(N1),color="data")) +
  geom_point(aes(y=ar3.fit.n1, color = "AR1 fit"), alpha=0.5) +
  geom_point(aes(y=lm3.fit.n1, color = "lm fit"), alpha=0.5) +
  geom_line(aes(y=log(N1), color="data"), linewidth=.4) +
  geom_line(aes(y=ar3.fit.n1, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
  geom_line(aes(y=lm3.fit.n1, color = "lm fit"), alpha=0.5, linewidth=.4) +
  theme_bw() +
  xlab("time") +
  ylab("N1") +
  labs(color="Legend") +
  scale_color_manual(values = colors)
p2 <- ggplot(data=dat,aes(x=time)) +
  geom_point(aes(y=log(N2),color="data")) +
  geom_point(aes(y=ar3.fit.n2, color = "AR1 fit"), alpha=0.5) +
  geom_point(aes(y=lm3.fit.n2, color = "lm fit"), alpha=0.5) +
  geom_line(aes(y=log(N2), color="data"), linewidth=.4) +
  geom_line(aes(y=ar3.fit.n2, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
  geom_line(aes(y=lm3.fit.n2, color = "lm fit"), alpha=0.5, linewidth=.4) +
  theme_bw() +
  xlab("time") +
  ylab("N2") +
  labs(color="Legend") +
  scale_color_manual(values = colors)
p1+p2
```

```{r figs14, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size of Species 1 and Species 2 in the previous time step $N_t$. Also shown is the change in population size of Species 1 as a function of $N_t$ for both species."}
dN1 <- rep(NA, (t-1))
for (i in 1:(t-1)) {
  dN1[i] <- log(N$N1[i+1])-log(N$N1[i])
}
d.dN1 <- as.data.frame(dN1)
d.dN1$Nt <- log(dat$N1[1:(t-1)])
d.dN1$Nt_plus_1 <- log(dat$N1[2:t])
p1 <- ggplot2::ggplot(d.dN1, aes(x=Nt,y=Nt_plus_1)) + geom_point()

d.dN1 <- as.data.frame(dN1)
d.dN1$Nt <- log(dat$N1[1:(t-1)])
d.dN1$Nt_plus_1 <- log(dat$N1[2:t])
d.dN1$N2 <- log(dat$N2[2:t])
p2 <- ggplot2::ggplot(d.dN1, aes(x=N2,y=Nt_plus_1)) + geom_point()
p3 <- ggplot2::ggplot(d.dN1, aes(Nt, dN1)) + geom_point()
p4 <- ggplot2::ggplot(d.dN1, aes(N2, dN1)) + geom_point()

p1+p2+p3+p4
```

We can see that the population size of Species 2 does not have a linear relationship with that of Species 1.

```{r}
df <- data.frame(cbind(log(dat$N1[2:t]),log(dat$N1[1:(t-1)]),log(dat$N2[1:(t-1)])))
colnames(df) <- c("Nt1","N1","N2")
m.3.lm <- lm(Nt1 ~ I(1/N1) + I(1/N2),data=df)

cor(df$Nt1, predict(m.3.lm))
summary(m.3.lm)
xx <- seq(0,100, length=1000)
prediction<-data.frame(N2=xx,N1=100)
plot(df$N2, df$Nt1, xlab = "N2", ylab="N.t+1" , pch=16)
lines(prediction$N2, predict(m.3.lm, prediction), lty=2,col="red",lwd=3)

```

```{r figs15, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size of the second species in the previous time step $N_t$. Fitted (line) and observed (points) values are shown."}
jtools::effect_plot(m.3.lm,pred=N2,plot.points = TRUE)
ggplot(data = df, aes(x = N2, y = Nt1)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x)
```
We can already see that the linear model does not provide a useful fit to the data points (neither does taking the inverse function help). The regression parameters can predict well the population size of the other species, but they are not informative for inference. This has been observed previously (@Certain2018), and non-linear least squares models (among others) are more often used to estimate parameters for dynamics resulting from these kinds of species interaction models (@Kloppers2013; @Muhlbauer2020; @Olivenca2021). The inability of the linear models to reproduce the interaction coefficients is more clearly demonstrated in @Certain2018. However, they do indicate that the slope of the linear model can reflect the direction of effect for the interacting species.

Instead of further exploring techniques to fit to time-series derived from non-linear competition processes, we look more closely at how HMSC manages to fit data emerging from competition dynamics and make inference about species interactions. Instead of estimating full interaction coefficients, or assuming sparse interactions, they instead assume that a reduced number of linear combinations of species abundances that are most relevant to determining future growth rates for species in the community can bypass the 'curse of dimensionality' problem. This is well presented in @Ovaskainen2017. We apply that approach here:

## Bayesian linear statistical model: HMSC
We can fit the data to a linear model using HMSC, with latent variables to capture the species associations. These are implemented by including a random effect at the time level.

```{r}
# prepare data in HMSC format
Y <- as.matrix(cbind(log(dat$N1[2:t]),log(dat$N2[2:t])))
XData <- df[,2:3]
studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)
m.3.hmsc = Hmsc(Y = Y, XData = XData, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 1000
transient <- 500*thin
verbose <- 500*thin
# sample MCMC
m.3.sample <- sampleMcmc(m.3.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
```

```{r}
m3.post.hmsc <- convertToCodaObject(m.3.sample)
summary(m3.post.hmsc$Beta)
bayesplot::mcmc_trace(m3.post.hmsc$Beta)
bayesplot::mcmc_areas(m3.post.hmsc$Beta,area_method = c("equal height"))
```

Now we look at the estimates for the species associations:

```{r}
OmegaCor = computeAssociations(m.3.sample)
supportLevel = 0.95
toPlot = ((OmegaCor[[1]]$support>supportLevel)
+ (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
corrplot::corrplot(toPlot, method = "color",col = colorRampPalette(c("blue","white","red"))(200),title = paste("random effect level:", m.3.sample$rLNames[1]), mar=c(0,0,1,0))
```

The model fixed effect estimates match well with those from the AR1 and linear model. However, the model up to this point includes $N_{t1}$ and $N_{t2}$ as fixed predictors. I change that here, as it better matches what can be done in large multi-species models, and the residual associations will better match the effects species have on one another.

```{r}
# AR1 coefficients (recall that the intercept is the term below multipled by 1 - phi1)
m.3.ar.n1$coef
m.3.ar.n1$coef[2]*(1-m.3.ar.n1$coef[1])
# linear model
summary(m.3.lm.n1)$coefficients[1:3,1:2]
# Bayesian estimates
summary(m3.post.hmsc$Beta)$statistics[1:3,1:2]
```

I try again with a next example where the only difference between species is the $\alpha_{ij}$.

```{r}
# Simulate initial species population growth with a second species present
N1.0 <- 10
N2.0 <- 10
r1.0 <- 1.7
r2.0 <- 1.7
alpha.11 <- 0.01
alpha.22 <- 0.01
alpha.12 <- 0.005
alpha.21 <- 0.01
# model function
disc_LV_comp <- function(r1,r2,N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21) {
    Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
    Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
    return(c(Nt1,Nt2))
}
# Simulation of model for t time steps
t <- 40
N <- array(NA,dim=c(t,2))
N <- as.data.frame(N)
colnames(N) <- c("N1","N2")
N$N1[1] <- N1.0
N$N2[1] <- N2.0
for (i in 2:t) {
  res <- disc_LV_comp(r1=r1.0,r2=r2.0,N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21)
  N$N1[i] <- res[1]
  N$N2[i] <- res[2]
}
```
```{r}
# Plot simulation: ggplot
N$time <- 1:t
dat <- melt(N, id.vars="time")
ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
```

```{r}
# Plot simulation: ggplot
dat <- as.data.frame(cbind(N$N1,N$N2))
colnames(dat) <- c("N1","N2")
dat$time <- 1:t
df <- data.frame(cbind(log(dat$N1[2:t]),log(dat$N2[2:t])))
colnames(df) <- c("Nt1","Nt2")
# prepare data in HMSC format
Y <- as.matrix(cbind(df$Nt1,df$Nt2))
XData <- data.frame(cbind(log(dat$N1[1:(t-1)]),log(dat$N2[1:(t-1)])))
studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)
m.4.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~1, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 1000
transient <- 500*thin
verbose <- 500*thin
# sample MCMC
m.4.sample <- sampleMcmc(m.4.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
```

```{r}
m4.post.hmsc <- convertToCodaObject(m.4.sample)
summary(m4.post.hmsc$Beta)
bayesplot::mcmc_trace(m4.post.hmsc$Beta)
bayesplot::mcmc_areas(m4.post.hmsc$Beta,area_method = c("equal height"))

OmegaCor = computeAssociations(m.4.sample)
supportLevel = 0.95
toPlot = ((OmegaCor[[1]]$support>supportLevel)
+ (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
corrplot::corrplot(toPlot, method = "color",col = colorRampPalette(c("blue","white","red"))(200),title = paste("random effect level:", m.4.sample$rLNames[1]), mar=c(0,0,1,0))
```

>>>>

---
title: "Appendix S2. Evolutionary drivers of species and community dynamics"
author:
  - Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]
  - Ruben J. Hermann^[University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen, Germany, ruben.hermann@uni-due.de]
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    extra_dependencies: ["float"]
  bookdown::html_document2: default
  word_document:
    reference_docx: docx_template.docx
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Author Pantel and Hermann}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
link-citations: yes
linkcolor: blue
csl: the-american-naturalist.csl
bibliography: refs.bib
---

@@

# Background

In Appendix S1.5, we evaluated the ability of the statistical model to estimate the impact of trait evolution for species abundances. Trait evolution's impact for species abundances can be modeled as a predictor in different ways. Here we derive the numerical form of the trait $x_i$ that we hypothesize has the most impact on species abundances.

# Model for population growth, competition, environmental change, and evolution

The growth equation for each species is:
